Forecasting the 2024 U.S. Presidential Election

### 2024-11-01 paper/paper.qmd
update result and discussion

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "Using Aggregated Polling Data to Predict the Outcome of the 2024 Presidential Race"
author: 
  - Andy Jiang
  - Onon Burentuvshin
  - Rui Hu
thanks: "Code and data are available at: [https://github.com/AndyYanxunJiang/president-polls)."
date: today
date-format: long
abstract: "The 2024 U.S. presidential election is shaping up to be a highly competitive race, with polling data playing a crucial role in gauging public sentiment and forecasting the potential outcome. This paper aims to build a predictive model using a poll-of-polls approach, aggregating data from multiple pollsters to forecast the winner of the upcoming election. The analysis incorporates polling data from various sources, filtered to include only high-quality pollsters with a numeric grade above 3, to ensure the reliability of the predictions. Additionally, a detailed examination of one selected pollster’s methodology is conducted, highlighting strengths, weaknesses, and biases in the survey process. The paper concludes with an idealized polling methodology, designed within a $100K budget, to enhance the accuracy of election forecasts. Results are presented in the context of understanding polling trends, potential biases, and the evolving landscape of voter preferences as the election approaches."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The analysis will further explore the geographic distribution of support across key battleground states, providing insights into the electoral landscape. Finally, the paper presents an idealized polling methodology with a $100,000 budget, designed to improve the accuracy and representativeness of election forecasts. By integrating multiple data sources and critically evaluating polling practices, this study aims to contribute to the ongoing discussion on the role of polling in democratic processes and election forecasting.

@@

sample_data <- data %>% sample_n(10)

# Display the sampled data as a table with smaller text
kable(sample_data, caption = "Sample of 10 Random Observations from the Analysis Data") %>%
  kable_styling(font_size = 8)  # Adjust the font size as needed
```

@@

# Results

```{r}
#| echo: false
#| warning: false
#| message: false
data_filtered <- data %>% filter(candidate_name %in% c("Donald Trump", "Kamala Harris"))

# Now proceed with calculating the average polling percentage by state
data_avg <- data_filtered %>%
  group_by(state, candidate_name) %>%
  summarize(avg_pct = mean(pct, na.rm = TRUE))

# Plotting the average polling percentages by state
ggplot(data_avg, aes(x = reorder(state, avg_pct), y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Polling Percentages by State",
       subtitle = "Comparing Support for Trump and Harris Across States",
       x = "State",
       y = "Average Polling Percentage",
       fill = "Candidate") +
  coord_flip() +  # Flip coordinates for easier reading
  theme_minimal()
```

@@

@@

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

@@

Method: Stratified sampling is used, where different voter groups are targeted based on their demographic and regional characteristics. For battleground state polls, the sample is drawn from states likely to decide the election, such as Arizona, North Carolina, Georgia, and Pennsylvania.

@@

Assumption 2:  Based on historical and polling data, some states’ electoral votes are predetermined as either Democrat or Republican. Therefore, it is more prudent to shift focus over to swing states where outcomes are uncertain. This assumption is accompanied with the challenge of determining what states qualify as “swing states”. However, on this issue, there is consensus on which states qualify as battleground states. As of 2024, our methodology should direct all our resources to acquiring voting data on the following states (New York Times).
Pennsylvania
Georgia
North Carolina
Michigan
Arizona
Wisconsin
Nevada

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election"
subtitle: "Using Aggregated Polling Data to Predict the Outcome of the 2024 Presidential Race"
author: 
  - Andy Jiang
  - Onon Burentuvshin
  - Rui Hu
thanks: "Code and data are available at: [https://github.com/AndyYanxunJiang/president-polls)."
date: today
date-format: long
abstract: "The 2024 U.S. presidential election is shaping up to be a highly swing race, with polling data playing a crucial role in gauging public sentiment and forecasting the potential outcome. This paper aims to build a predictive model using a poll-of-polls approach, aggregating data from multiple pollsters to forecast the winner of the upcoming election. The analysis incorporates polling data from various sources, filtered to include only high-quality pollsters with a numeric grade above 3, to ensure the reliability of the predictions. Additionally, a detailed examination of one selected pollster’s methodology is conducted, highlighting strengths, weaknesses, and biases in the survey process. The paper concludes with an idealized polling methodology, designed within a $100K budget, to enhance the accuracy of election forecasts. Results are presented in the context of understanding polling trends, potential biases, and the evolving landscape of voter preferences as the election approaches."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The analysis will further explore the geographic distribution of support across key swing states, providing insights into the electoral landscape. Finally, the paper presents an idealized polling methodology with a $100,000 budget, designed to improve the accuracy and representativeness of election forecasts. By integrating multiple data sources and critically evaluating polling practices, this study aims to contribute to the ongoing discussion on the role of polling in democratic processes and election forecasting.

@@

# Sample 10 random observations from the filtered data with Trump and Harris
sample_data <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  sample_n(10)

# Display the sampled data as a table with smaller text and adjusted column spacing
kable(sample_data, caption = "Sample of 10 Random Observations from the Analysis Data") %>%
  kable_styling(font_size = 8, full_width = FALSE) %>%  # Set full_width to FALSE to center the table
  column_spec(1:ncol(sample_data), width = "2cm")  # Adjust width as needed, here set to 2cm per column

@@

# Results
The analysis of polling data across key swing states for the 2024 U.S. Presidential Election reveals insights into the levels of support for candidates Donald Trump and Kamala Harris. The findings are presented through three primary visualizations: the average polling percentage by state, the distribution of polling percentages, and the difference in average polling percentage between the two candidates.

## Average Polling Percentage by State
From @fig-1, we can see that in the swing states of Arizona, Georgia, and North Carolina, Trump holds a narrow lead. In contrast, Harris leads in Pennsylvania, Wisconsin, Nevada, and Michigan. When looking at the overall average across these states, Harris slightly edges out Trump, indicating her overall advantage in these swing areas.

## Polling Percentage Distribution
The box plot in @fig-2 demonstrates the range and consistency of support for both candidates across each state. Trump shows a wider range of support, reflecting variability, while Harris has relatively steady polling percentages across states, maintaining higher consistency. Harris’s support is less variable and generally centers around higher polling percentages compared to Trump’s.

## Difference in Average Polling Percentage
@fig-3 highlights the difference in average polling percentages by state, with positive values indicating Trump’s lead and negative values indicating Harris's. Trump’s lead in the three states (Arizona, Georgia, and North Carolina) is relatively narrow, with margins of less than 1 percent. In contrast, Harris’s lead in the states favoring her (Pennsylvania, Wisconsin, Nevada, and Michigan) ranges from 1 to 4 percent, indicating her stronger support in the states where she leads.

```{r}
#| label: fig-1
#| fig-cap: "Average Polling Percentage by State"
#| echo: false
#| warning: false
#| message: false
# Calculate the average polling percentage by state for Trump and Harris
data_avg <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  group_by(state, candidate_name) %>%
  summarize(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

# Calculate the overall average polling percentage for Trump and Harris
overall_avg <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  group_by(candidate_name) %>%
  summarize(avg_pct = mean(pct, na.rm = TRUE)) %>%
  mutate(state = " Overall")  # Add a space before "Overall" to separate it visually

# Combine state-level data with overall average
combined_data <- bind_rows(data_avg, overall_avg)

# Ensure "Overall" appears at the bottom with spacing
combined_data$state <- factor(combined_data$state, levels = c(setdiff(unique(combined_data$state), " Overall"), " Overall"))

# Plot the average polling percentages by state with overall comparison
ggplot(combined_data, aes(x = state, y = avg_pct, fill = candidate_name)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), aes(color = ifelse(state == " Overall", "Overall", "State")), width = 0.6) + 
  scale_color_manual(values = c("Overall" = "black", "State" = NA), guide = "none") +  # Add dark grey outline for "Overall" bars
  scale_fill_manual(values = c("Donald Trump" = "#F8766D", "Kamala Harris" = "#56B4E9")) +  # Custom colors for consistency
  labs(title = "Average Polling Percentages by State",
       x = "State",
       y = "Average Polling Percentage",
       fill = "Candidate") +
  coord_flip() +  # Flip coordinates for easier reading
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(margin = margin(r = 10)), # Add spacing to y-axis labels
    panel.grid.major.y = element_blank(), # Remove y-axis gridlines for clarity
    panel.spacing = unit(1, "lines") # Increase panel spacing
  )

@@

```{r}
#| label: fig-2
#| fig-cap: "Polling Percentage Distribution"
#| echo: false
#| warning: false
#| message: false
ggplot(data %>% filter(candidate_name %in% c("Donald Trump", "Kamala Harris")), 
       aes(x = state, y = pct, fill = candidate_name)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Polling Percentage Distribution by State",
       x = "State",
       y = "Polling Percentage",
       fill = "Candidate") +
  scale_fill_manual(values = c("Donald Trump" = "#F8766D", "Kamala Harris" = "#56B4E9")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
```{r}
#| label: fig-3
#| fig-cap: "Difference in Average Polling Percentage"
#| echo: false
#| warning: false
#| message: false
data_diff <- data %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) %>%
  group_by(state) %>%
  summarize(diff_pct = mean(pct[candidate_name == "Donald Trump"], na.rm = TRUE) - 
                      mean(pct[candidate_name == "Kamala Harris"], na.rm = TRUE), .groups = "drop")

ggplot(data_diff, aes(x = reorder(state, diff_pct), y = diff_pct, fill = diff_pct > 0)) +
  geom_bar(stat = "identity", color = "darkgrey") +
  scale_fill_manual(values = c("TRUE" = "#F8766D", "FALSE" = "#56B4E9"), 
                    labels = c("Kamala Harris", "Donald Trump")) +
  labs(title = "Difference in Average Polling Percentage by State",
       x = "State",
       y = "Difference in Polling Percentage",
       fill = "Lead") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

## Impact of Key Demographics and Policy Appeals
Harris’s lead in several swing states may be influenced by her campaign’s appeal to key demographic groups, particularly immigrants. The Democratic platform has emphasized social and economic benefits for immigrant communities, which may have strengthened her base in states with significant immigrant populations. Policies that promise support for healthcare, education, and other public services are likely contributing to Harris’s appeal among these groups, solidifying her lead in states like Nevada and Michigan. This demographic support provides Harris with a strategic advantage, as these communities may play a crucial role in voter turnout and loyalty.

## Influence of Endorsements and High-Profile Backing
Recently, high-profile figures like Elon Musk have publicly backed Trump, using their influence to mobilize support for his campaign. Musk’s endorsement may energize Trump’s voter base, especially among his followers and individuals interested in business and technology. This kind of visible support from influential figures could affect last-minute voter decisions, particularly in tightly contested states where Trump’s lead is slim. If Musk’s endorsement translates into increased campaign funding or social media influence, it could bolster Trump’s standing in swing states where the margin remains close, potentially altering the final outcome on November 5th.

## Limitations of the Analysis: Focus on Swing States
While the analysis focuses on key swing states, it does not consider the broader national picture, which includes states with clear, one-sided advantages for either candidate. This narrowed focus on swing states provides insights into swing areas but does not give a complete picture of the national landscape. Therefore, while Harris appears to have an edge in these swing states, it is unclear whether she is leading overall. The situation in solidly red or blue states could still shift the electoral balance, meaning Trump could potentially gain ground through strong performances in traditionally Republican states.

## Potential Influence of Late Campaign Efforts
With a few days remaining before the election, both campaigns are likely to intensify their outreach efforts, particularly in swing states. Trump’s recent increase in targeted advertisements and campaign rallies in swing states could energize his supporters, narrowing Harris’s lead. On the other hand, Harris’s campaign may focus on mobilizing key demographics, such as young and minority voters, to secure her advantage. Late-breaking events, public debates, or controversies could sway undecided voters and ultimately impact the election results in these states.

## Limitations and Next Steps
This analysis is limited to examining current polling data in swing states and does not account for unexpected shifts in voter sentiment or turnout on election day. Additionally, focusing exclusively on swing states omits the influence of non-swing states that may still play a role in the electoral vote tally. Future analyses could incorporate national polling data to provide a more comprehensive perspective and examine demographic breakdowns within each state to understand which groups are driving support for each candidate. Integrating data on recent campaign activities, endorsements, and voter sentiment on pressing issues could also improve the predictive accuracy of election outcomes.

@@

Method: Stratified sampling is used, where different voter groups are targeted based on their demographic and regional characteristics. For swing state polls, the sample is drawn from states likely to decide the election, such as Arizona, North Carolina, Georgia, and Pennsylvania.

@@

Assumption 2:  Based on historical and polling data, some states’ electoral votes are predetermined as either Democrat or Republican. Therefore, it is more prudent to shift focus over to swing states where outcomes are uncertain. This assumption is accompanied with the challenge of determining what states qualify as “swing states”. However, on this issue, there is consensus on which states qualify as swing states. As of 2024, our methodology should direct all our resources to acquiring voting data on the following states (New York Times).
Pennsylvania
Georgia
North Carolina
Michigan
Arizona
Wisconsin
Nevada
